Draft 2

### 2020-06-30 Paper_V2.Rmd
Correcting code... adding time series

---
title: "Draft 2"
author: "Ana Karina"
date: "4/7/2020"
output: html_document
bibliography: references.bib
---

@@

### Exposure Assessment 
#### Scenario definition
Only one potential exposure pathway was analysed: the indirect transmission of viruses through contaminated surfaces. Two different scenarios where modelled: single touch (frequently touched buttons) and multiple touches (ATM or cash machine).

@@

####  Virus concentration and transfer 
Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input. The virus titers used in the risk assessment were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020]. The RT-qPCR method estimates the number of copies of viral RNA present in the samples. Nevertheless, the presence of coronavirus RNA does not imply the presence of infectious viruses. In risk assessments whenever data on the concentration of infectious viruses is not available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses (REF). The gene copy number to infectivity ratio is, to our knowledge, not available for SARS-CoV-2. In the absence of SARS-CoV-2 data, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016 (Table 1). Both, influenza viruses and coronaviruses, are enveloped, RNA viruses with similar sizes.

@@

#### Model parameters used in the risk assessment 
| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
|:--:|:--:|:--:|:--:|:--:|:--:|
| $C_{spGC}$ | gene copies/mL | Concentration of SARS-CoV-2 in sputum or saliva | Log~10~ gene copies/mL (data set) | ReSample | [@Wolfel2020; @Pan2020; @Kim2020; @To2020] RT-qPCR data from 9, 2, 2, and 23 patients with COVID-19|
| $GC:TCID_{50}$ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | [@Ip2016] Mean RT-qPCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
| $C_{sp}$ | TCID~50~/mL |Concentration of infective SARS-CoV-2 in sample | $C_{sp}=\frac {C_{spGC}}{GC:TCID_{50}}$ | - | calculated|
| $V_s$ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | [@Nicas2009] volume = 0.044mL, [@Adhikari2019] uniform $\pm$ 10%|
|$t_{ATM}$ | min |Time spent on the ATM | 0.75(0.38) | Normal | This study. Otained from 20 hours of observational data |
|$t_{btw-ATM}$ | min |Time between visits | 8.4(7.1) | Normal | This study. Otained from 20 hours of observational data |
|$t_{btw-btn}$ | min |Time between pressing button | 0.5-20 | Uniform | Assumed |
|$x_{ATM}$| cm | ATM inoculation distance | 40 | Point value | This study. Assumed from the observational data  |
|$x_{btn}$| cm | Button inoculation distance | 50 | Point value | Assumed |
|$t_{\frac {1}{2}stl}$| min | Half life of CoV-2 in steel | 338(35) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
|$t_{\frac {1}{2}pl}$| min | Half life of CoV-2 in plastic | 409(39) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
| $n_x$| min^-1^ |Decay rate in surface $x$ (steel or plastic) | $n_x=\frac {ln(2)}{t_\frac{1}{2}}$ | - | calculated |
| $TE_{sh-stl}$ | % | Transfer efficiency of virus between steel to hand | 37.4(16) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
| $TE_{sh-pl}$ | % | Transfer efficiency of virus between plastic to hand | 79.5(21.2) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
| $TE_{hm}$ | % | Transfer efficiency from hand to mucouse membranes | 20(6.3) | Normal | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
| $A_{f}$ | cm^2^ | Finger surface area | 3.8-6.6 | Uniform | Data from @Chabrelie2018 @ @Peters2009a @Sahmel2015 | 
| $FSA_{fm}$ | - | Fractional surface area between finger and mucous membranes | 0.5-0.8 | Uniform | Assumed | 
| $k$ | PFU^-1^ | Parameter of the exponential dose-response| 0.00107, 0.00135, 0.00680 | Triangular | Data obtained from QMRAwiki, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 0.5th, 50th, and 99.5th percentiles as min, mode, and max |
|:--:|:--:|:--:|:--:|:--:|:--:|
| N~cgh/min~ |min^-1^ | Number of coughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV|

### Risk Characterization 
Monte Carlo simulation method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.

>>>>

---
title: "Draft 2"
author: "Ana Karina"
date: "4/7/2020"
#output: html_document
output: pdf_document
bibliography: references.bib
---

@@

### Exposure Assessment 
#### Scenario definition
Only one potential exposure pathway was analysed: the indirect transmission of viruses through contaminated surfaces. Two different scenarios where modelled: single touch (frequently touched buttons) and multiple touches (ATM or cash machine). The stochastic and mechanistic model used to model the risk of surface-mediated transmission of SARS-CoV-2 infection is based on models described elsewhere [@Julian2018; @Julian2019].

@@

####  Virus concentration and transfer 
Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input. The virus titers used in the risk assessment were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020]. The RT-qPCR method estimates the number of copies of viral RNA present in the samples. Nevertheless, the presence of coronavirus RNA does not imply the presence of infectious viruses. In risk assessments whenever data on the concentration of infectious viruses is not available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses [@Chigor2014]. The gene copy number to infectivity ratio is, to our knowledge, not available for SARS-CoV-2. In the absence of SARS-CoV-2 data, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016 (Table 1). Both, influenza viruses and coronaviruses, are enveloped, RNA viruses with similar sizes.

@@

#### Model parameters used in the risk assessment 
| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
|:--:|:--:|:--:|:--:|:--:|:--:|
| $C_{spGC}$ | gene copies/mL | Concentration of SARS-CoV-2 in sputum or saliva | Log~10~ gene copies/mL (data set) | ReSample | [@Wolfel2020; @Pan2020; @Kim2020; @To2020] RT-qPCR data from 9, 2, 2, and 23 patients with COVID-19|
| $GC:TCID_{50}$ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | [@Ip2016] Mean RT-qPCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
| $C_{sp}$ | TCID~50~/mL |Concentration of infective SARS-CoV-2 in sample | $C_{sp}=\frac {C_{spGC}}{GC:TCID_{50}}$ | - | calculated|
| $V_s$ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | [@Nicas2009] volume = 0.044mL, [@Adhikari2019] uniform $\pm$ 10%|
| $N_{cgh/min}$ |min^-1^ | Number of coughs per minute  | 0.57(1)| Normal |[@Leung2020], based on data of 17 individuals with seasonal CoV| 
|$t_{ATM}$ | min |Time spent on the ATM | 0.75(0.38) | Normal | This study. Otained from 20 hours of observational data |
|$t_{btw-ATM}$ | min |Time between visits | 0.13, 0.80 | Weibull | This study. Otained from 20 hours of observational data |
|$t_{btw-btn}$ | min |Time between pressing button | 1-20 | Uniform | Assumed |
|$x_{ATM}$| cm | ATM inoculation distance | 40 | Point value | This study. Assumed from the observational data  |
|$x_{btn}$| cm | Button inoculation distance | 50 | Point value | Assumed |
|$t_{\frac {1}{2}stl}$| min | Half life of CoV-2 in steel | 338(35) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
|$t_{\frac {1}{2}pl}$| min | Half life of CoV-2 in plastic | 409(39) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
| $n_x$| min^-1^ |Decay rate in surface $x$ (steel or plastic) | $n_x=\frac {ln(2)}{t_\frac{1}{2}}$ | - | calculated |
| $TE_{sh-stl}$ | % | Transfer efficiency of virus between steel to hand | 37.4(16) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
| $TE_{sh-pl}$ | % | Transfer efficiency of virus between plastic to hand | 79.5(21.2) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
| $TE_{hm}$ | % | Transfer efficiency from hand to mucouse membranes | 20(6.3) | Normal | [@Pitol2017], transfer efficiency of viruses (MS2) from hand to saliva | 
| $A_{f}$ | cm^2^ | Finger surface area | 3.9-5.9 | Uniform | Fractional surface area of partial finger [@AuYeung2008] times average hand surface area [@EPA2011]| 
| $k$ | PFU^-1^ | Parameter of the exponential dose-response| 0.00107, 0.00135, 0.00680 | Triangular | Data obtained from QMRAwiki, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 0.5th, 50th, and 99.5th percentiles as min, mode, and max |
|:--:|:--:|:--:|:--:|:--:|:--:|
The input parameters of the distributions are: Normal: mean (SD); Uniform: lower-bound - upper-bound; Triangular: min, mode, max; Weibull: scale, shape.

### Risk Characterization 
Monte Carlo simulation: method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.
